In-Band Synchronization for\\Distributed SDN Control Planes\thanks{Supported by the European Research Council (ERC) Starting Grant no.~259085, the European Institute of Innovation \& Technology (EIT) project 13153 (Software Defined Networking) and the FP7 EU project UNIFY.}

### 2015-11-20 ccr-camera/inband-sync.tex
final version

%[[PK sorry for stubborness :)
%\title{On the Synchronization Power of the OpenFlow Data Plane}
\title{In-Band Synchronization for\\Distributed SDN Control Planes\thanks{Supported by the
European Institute of Innovation \& Technology (EIT) project 13153 (Software Defined Networking) 
and the FP7 EU project UNIFY.}}
%\\ {\Large Towards Atomic Transactions In-Band}}
%]]

@@

\begin{abstract}
Control planes of forthcoming Software-Defined Networks (SDNs)
will be \emph{distributed}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
modules of the control plane %operating on the data plane 
should be physically distributed.
However, %a distributed control plane also introduces
%new challenges,
in order to guarantee \emph{consistency} of network operation, actions
performed on the data plane by different controllers may need to 
be \emph{synchronized},  which is a
 nontrivial task.
In this paper, we propose a synchronization
framework for control planes  based on atomic
transactions, 
%In particular, we make the case for 
implemented \emph{in-band}, on the 
data-plane switches.
We argue that this in-band approach is attractive as 
it keeps the failure scope local and 
does not require additional 
out-of-band coordination mechanisms.
It allows us to realize fundamental consensus primitives in the presence
of controller failures, and we discuss their applications
for consistent policy composition and fault-tolerant control-planes.
Interestingly, 
by using part of
the data plane configuration space as a shared memory
and leveraging  the match-action
paradigm, 
%[[PK key-valued stores are not really discussed?
%to realize simple key-value stores, 
we can implement our synchronization framework in
today's standard OpenFlow protocol, 
and we report on
our proof-of-concept implementation.
\end{abstract}

@@

To check whether a value $k$ is written at a given
address $\addr$ (Algorithm ~\ref{alg:compare}),
a controller tries to add a new entry the same way as in the \memwrite operation,
but additionally setting the \texttt{\checko} flag. 
If a different value $k'$ is currently written, 
%our action differs from the existing entry, but 
the entry inflicts a failure, as it overlaps but does not coincide
with the existing entry.
Indeed, the value parts of the two entries are identical, but the
masks differ.
% \liron{(different mixtures of ones and wildcards in the $k'$ part)},
%thereby inflicting a failure. 
In case the currently written value is $k$, the existing flow entry is replaced
with an identical one.
%[[PK on initialization
Before using an address $\addr$, it is advised that a controller safely
initializes it with a special initial value $0$ by calling
$\textit{\compare}(\addr,0)$: 
the first such command succeeds, and any 
subsequent one will fail. 
%]]

@@

Let us now revisit our load-balancing application
discussed in Section~\ref{sec:motivation}.
A simple strategy to balance load in our 2-link example
 (Figure~\ref{fig:CAS-example}) fails in the presence of
multiple controllers, even if 
bundling and $\checko$ features are used:
the flows are defined over ``independent'' 
flow spaces.

@@

Algorithm~\ref{alg:simple-update} is correct assuming that policy
identifiers may grow without bound. 
In practice, we can use a modulo scheme on the 32-bit field of a flow entry
used for storing the policy identifiers, which, in principle,
may lead to inconsistencies if controllers proceed in different speeds.
Algorithm~\ref{alg:update} adresses this issue 
by using our id-claimer abstraction.
%to bound the number of used policy identifiers.
Here, the controller first reads the current policy id and 
claims it, which prevents another controller from using the id 
for a different policy
(Lines~\ref{update:read-state1}-\ref{update:read-state2}). 
Then the controller selects any \emph{unused} id and computes the update commands to be executed (Line~\ref{update:compute}). %\liron{I have replaced "claims the first" with "chooses an" following Petr email}
Finally, within an atomic transaction (\textbf{\execatomic}), it verifies whether the chosen policy id has not
been claimed and that
the policy has not been concurrently modified, and then tries to update the
policy.
%

>>>>

%[[PK sorry for stubborness :)
%\title{On the Synchronization Power of the OpenFlow Data Plane}
\title{In-Band Synchronization for\\Distributed SDN Control Planes\thanks{Supported by the European Research Council (ERC) Starting Grant no.~259085, the
European Institute of Innovation \& Technology (EIT) project 13153 (Software Defined Networking) 
and the FP7 EU project UNIFY.}}
%\\ {\Large Towards Atomic Transactions In-Band}}
%]]

@@

\begin{abstract}
Control planes of forthcoming Software-Defined Networks (SDNs)
will be \emph{distributed}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
modules of the control plane 
should be physically distributed.
However, 
in order to guarantee \emph{consistency} of network operation, actions
performed on the data plane by different controllers may need to 
be \emph{synchronized},  which is a
 nontrivial task.
In this paper, we propose a synchronization
framework for control planes  based on atomic
transactions, 
implemented \emph{in-band}, on the 
data-plane switches.
We argue that this in-band approach is attractive as 
it keeps the failure scope local and 
does not require additional 
out-of-band coordination mechanisms.
It allows us to realize fundamental consensus primitives in the presence
of controller failures, and we discuss their applications
for consistent policy composition and fault-tolerant control-planes.
Interestingly, 
by using part of
the data plane configuration space as a shared memory
and leveraging  the match-action
paradigm, 
we can implement our synchronization framework in
today's standard OpenFlow protocol, 
and we report on
our proof-of-concept implementation.
\end{abstract}

@@

To check whether a value $k$ is written at a given
address $\addr$ (Algorithm ~\ref{alg:compare}),
a controller tries to add a new entry the same way as in the \memwrite operation,
but additionally setting the \texttt{\checko} flag. 
If a different value $k'$ is currently written, 
%our action differs from the existing entry, but 
the entry inflicts a failure, as it overlaps but does not coincide
with the existing entry.
Indeed, the value parts of the two entries are identical, but the
masks differ.
% \liron{(different mixtures of ones and wildcards in the $k'$ part)},
%thereby inflicting a failure. 
In case the currently written value is $k$, the existing flow entry is replaced
with an identical one.
%[[PK on initialization
Before using an address $\addr$, it is advised that a controller safely
initializes it with a special initial value $0$ by calling
$\textit{\compare}(\addr,0)$: 
the first such command succeeds, 
and subsequent ones will do no harm.
%]]

@@

Let us now revisit our load-balancing application
discussed in Figure~\ref{fig:CAS-example}.
A simple strategy to balance load in our 2-link example
 (Figure~\ref{fig:CAS-example}) fails in the presence of
multiple controllers, even if 
bundling and $\checko$ features are used:
the flows are defined over ``independent'' 
flow spaces.

@@

Algorithm~\ref{alg:simple-update} is correct assuming that policy
identifiers may grow without bound. 
In practice, we can use a modulo scheme on the 32-bit field of a flow entry
used for storing the policy identifiers, which, in principle,
may lead to inconsistencies if controllers proceed in different speeds.
Algorithm~\ref{alg:update} adresses this issue 
by using our id-claimer abstraction.
%to bound the number of used policy identifiers.
Here, the controller first reads the current policy id and 
claims it, which prevents another controller from using the id 
for a different policy
(Lines~\ref{update:read-state1}-\ref{update:read-state2}). 
Then the controller chooses any \emph{unused} id and computes the update commands to be executed (Line~\ref{update:compute}). %\liron{I have replaced "claims the first" with "chooses an" following Petr email}
Finally, within an atomic transaction (\textbf{\execatomic}), it verifies whether the chosen policy id has not
been claimed and that
the policy has not been concurrently modified, and then tries to update the
policy.
%
